Loan Analysis (Prosper Provided) by Andrew Teixeira
========================================================

#Initial Data Loading and Exploration 

The data, provided by Prosper and made available by Udacity through the 
Data Analyst program, is data on loans made through the Prosper platform,
an online fintech company that offers peer-to-peer financing in which 
individuals can loan money to individual borrowers for specific purposes,
a different means of financing from the traditional loans. 

In the dataset, we can examine and explore the characteristics of a 
Prosper loan (Interest Rate, APR etc.) and a Prosper borrower 
(Credit Score etc.) that may make for an interesting insight into how the 
platform operates, finances itself, and is lucrative for investors. 


```{r}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


##Load all relevant libraries


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(plyr)
library(gridExtra)
library(knitr)
```

##Establish working directory
```{r}
#Establishing the working directory
getwd()
setwd('/Users/andrewteixeira/Documents/DAND/Final_Project_EDA')
getwd()
```

Load the data into the variable "loans"
```{r echo=FALSE, Load_the_Data}
loans <- read.csv('prosperLoanData.csv')
```

###Examining the column names

```{r}
names(loans)
head(loans)
```


# Univariate Plots Section

We will be looking at a few variables that relate to both the loan itself 
(dollar amount, term) and the borrower (credit score, prosper rating) in order 
to determine what is likely to influence the total amount Prosper is willing to 
lend a borrower and at what rate. 

##Borrower Data Exploration: Subsection: Loan Original Amount

We will be examining the LoanOriginalAmount data from the loans dataset to 
better understand the frequency of loan values provided by Prosper. 

```{r}
head(sort(table(loans$LoanOriginalAmount), decreasing = T), 20)
```

```{r}
summary(loans$LoanOriginalAmount)
```

The table above shows that loans of 4,000 are the most provided by Propser. 
This value is below the median of 6500 and the Mean of 8337, which suggests 
there are some large loans that are pulling the mean forward.

The max loan provided by Prosper was 35,000.

```{r}
qplot(data = loans, x = loans$LoanOriginalAmount, binwidth = 1000)+
  geom_histogram()+
  scale_x_continuous(breaks = seq(0,35000, 2000))+
  ggtitle('Count of Loan Original Amount')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


The histogram above shows a spike at around 3,000- 4,000. This suggests there 
are a great number of loans provided in this area. Let's examine further. 

```{r}
qplot(data = loans, x = loans$LoanOriginalAmount, binwidth = 500)+
  geom_histogram()+
  coord_cartesian(xlim = c(0,5000))+
  scale_x_continuous(breaks = seq(0,5000, 1000))+
  ggtitle('Count of Loan Original Amount by Binwidth = 500')
```


```{r}
head(sort(table(loans$LoanOriginalAmount), decreasing = T), 20)
```


##Borrower Data Exploration: Subsection: Loan Term

Let's look at the loan term provided by Prosper to get a sense of what types of 
loans and duration of loans Prosper gives to their borrowers. 

```{r}
sort(table(loans$Term))
```


```{r}
table(loans$Term)
loans$Term <- factor(loans$Term, levels=c(12,36,60), ordered=TRUE)
ggplot(data = loans, aes(x = loans$Term), binwidth = 1)+
  geom_bar()+
  ggtitle('Count of loan terms')
```


From the table and histogram above, we see that Prosper only gives loans in 12,
36, and 60 month terms. The majority of loans are provided on a 36 month term 
payment plan. 


##Borrower Data Exploration: Subsection: Amount Delinquent


We will examine the amount of money that is delinquent per borrower per Prosper 
loan. 

```{r}
DelinquenciesValues <- subset(loans, loans$AmountDelinquent != 0)
tail(sort(table(DelinquenciesValues$AmountDelinquent)),20)
ggplot(data = DelinquenciesValues, aes(x = AmountDelinquent))+
  geom_histogram()+
  ggtitle('Count of Amount Delinquent')
```

```{r}
summary(loans$AmountDelinquent)
```

```{r}
head(sort(table(loans$AmountDelinquent), decreasing = TRUE), 50)
```

```{r}
DelinquenciesValues <- subset(loans, loans$AmountDelinquent != 0)
ggplot(data = DelinquenciesValues, aes(x = AmountDelinquent))+
  geom_histogram(binwidth = 10)+
  coord_cartesian(xlim = c(0,2000))+
  ggtitle('Count of Loans by Amount Delinquent')
```


The data above in the histograms and table shows that the majority of Prosper 
borrowers have 0 dollars in delinquency, a good sign if you are a lender. 


##Borrower Data Exploration: Subsection: Delinquencies Last 7 Years

Examining the Delinquencies Last 7 Years variable will help us 
understand the type of borrowers that are included in this dataset. 
DelinquenciesLast7Years may provide a glimpse into the 
reliability of borrowers in this set. 

```{r}
head(sort(table(loans$DelinquenciesLast7Years), decreasing = T), 20)
summary(loans$DelinquenciesLast7Years)
```


**Findings** : Unsurprisingly, we see the vast majority of delinquent loans in 
the last 7 years for borrowers at 0. The median reflects this by also being 0, 
however, a mean of 4.155 shows that there are some potentially large amounts of 
delinquencies in the last 7 years per borrowers that may be driving this 
value up. 

In the table, we see that there are, in fact, 110 borrowers who have 99 
delinquencies in the past 7 years. This appears to be a strange lift from 98 
(3 total borrowers). Is this a maximum value? Could it be human error? 

```{r}
qplot(data = loans ,x=DelinquenciesLast7Years, binwidth = 1)+
  geom_histogram()+
  scale_x_continuous()+
  scale_y_continuous(
    limits = c(0,100000)
  )+
  ggtitle('Count of Delinquencies Last 7 Years')
```


**Findings**: A simple histogram shows that, in fact, the majority of 
delinquencies in last 7 years for borrowers were negatively skewed, 
unsurprisingly considering that banks and lenders are in the business of 
ensuring their loans are paid back. Let's examine and plot that surreptitious 
the DelinquenciesLast7Years on the larger end of the plot.

```{r}
qplot(data = loans, x = loans$DelinquenciesLast7Years)+
  geom_histogram(binwidth = 1)+
  scale_x_continuous(
    limits = c(80,100)
  )+
  ggtitle('Count of Delinquencies Last 7 years')
  
```


**Findings** The histogram from DelinquenciesLast7Years > 79 show that the #99 
is a strange occurance that may be artifically driving the mean up for the total 
DelinquenciesLast7Years. We will attempt to determine, later in this analysis, 
what makes an individual become delinquent with some multivariate tests and 
plots. 

##Borrower Data Exploration: Subsection: BorrowerAPR
We are getting a better sense of the types of borrowers Prosper loans money to. 
Now, we will examine the BorrowerAPR to determine how Prosper makes $ of 
their customers. 

```{r}
qplot(data = loans ,x=BorrowerAPR)+
  geom_histogram(binwidth = .01)+
  ggtitle('Count of Loans by Borrower APR')
```


The findings above dictate a few peaks: Once between .15 and .20, again at .3, 
and once more at around .36. 

Let's create a sorted table to examine these trends. 

```{r}
head(sort(table(loans$BorrowerAPR), decreasing = T), 20)
```

```{r}
summary(loans$BorrowerAPR)
```


Interestingly, we see that Prosper's most commonly issued APR is .3672, this is 
well above the mean of .21976 and the median of .20976. 

Let's remove all NA's from APR for analysis - adding the Mean line, we can 
examine how many values are falling above or below the mean. 

```{r}
Clean_APR <- subset(loans, !is.na(BorrowerAPR))
mean(Clean_APR$BorrowerAPR)

qplot(data = Clean_APR ,x=BorrowerAPR, binwidth = .01)+
  geom_histogram()+
  scale_x_continuous(breaks = seq(0,0.4,.02))+
  geom_vline(xintercept = mean(Clean_APR$BorrowerAPR), color = 'green', 
             linetype = 'longdash')+
  ggtitle('Count of Loans by Borrower APR')
```


We see a number of peaks and valleys within the BorrowerAPR data, 
which suggests there are certain integers that Prosper commonly issues 
when it comes to BorrowerAPR. 

##Borrower Data Exploration: Subsection: Income Range

```{r}
summary(loans$IncomeRange)
```

```{r}
loans_IncomeRanges <- subset(loans, loans$IncomeRange != "Not displayed")
summary(loans_IncomeRanges$IncomeRange)
qplot(data=loans_IncomeRanges, x=loans_IncomeRanges$IncomeRange)+
  geom_bar()+
  ggtitle('Count of Loans by Income Range')
```


**Findings** 
Overall, the income of borrowers is failing between 25,000 and 74,999. 

```{r}
sort(table(loans$BorrowerState), decreasing = T)
```


From the table above, we can see that California has the largeset amount of 
users on Prosper with 14,717 borrowers, while North Dakota has the least with 
52 borrowers. 


##Borrower Subsection: TotalInquiries 

```{r}
qplot(data=loans, x=TotalInquiries)+
  geom_histogram()+
  ggtitle('Count of Loans by Total Inquiries')
```


```{r}
head(sort(table(loans$TotalInquiries), decreasing = T), 20)
```


The data above show that most borrowers have very few inquiries on their credit, 
potentially showing that Prosper selects those who are steadfast and committed 
to paying down their debts. Let's create a plot in which the number 0 is 
removed, showing only borrowers with 0 inquiries

```{r}
TotalInquiriesValues <- subset(loans, TotalInquiries != 0)
summary(TotalInquiriesValues$TotalInquiries)
qplot(data = TotalInquiriesValues, x=TotalInquiriesValues$TotalInquiries,
      geom = "histogram", binwidth = 1.0)+
  coord_cartesian(xlim=c(0,40))+
  ggtitle('Count of Total Inquiries > 0')+
  xlab('Total Inquiries')
```

From the plot above, we see a steep decline in total inquiries values after the
initial 1,2 or 3 total inquiries. 


```{r}
qplot(data=loans, x=DebtToIncomeRatio)+
  geom_histogram()+
  ggtitle('Count of Loans by Debt To Income Ratio')
```


Given that the 10 value appears to be an anomaly, let's run a summary on 
Debt to Income Ratio

```{r}
summary(loans$DebtToIncomeRatio)
```

Let's further examine Debt to Income Ratios that are above the 3rd Quartile.

```{r}
High_DTIR <- subset(loans, loans$DebtToIncomeRatio > .32)
sort(table(High_DTIR$DebtToIncomeRatio),decreasing = T)
```


**Findings** From the data above, that there are a few loans at various price 
points that have obscenely high debt to income ratios. Perhaps these folks have 
a great deal of collateral to put up. Let's look at all the DTI ratios under
or equal to 1. 

```{r}
DTI_Ratios <- subset(loans, loans$DebtToIncomeRatio <= 1.0)
qplot(data = DTI_Ratios, DTI_Ratios$DebtToIncomeRatio, geom = "histogram",
      bindwith = .10)+
  scale_x_continuous(breaks = seq(0,1.0, .10))+
  ggtitle('Debt-to-Income Ratios Less Than or Equal to 1.0')+
  xlab('Debt-to-Income Ratio')
```

From the chart above, we see that the majority of Debt-to-Income ratios are
falling under .5, no surprise as the 3rd quartile value is .32.

```{r}
qplot(data=loans, x=EmploymentStatus)+
  geom_bar()+
  ggtitle('Count of Loans by Employment Status')
```


The graph above shows that most borrowers are "employed" or categorized as 
"full-time"

```{r}
qplot(data = loans, x=ProsperRating..numeric.)+
  geom_bar()+
  ggtitle('Count of Loans by Prosper Rating Numeric')
```

Let's examine the credit scores of applicant's by creating a new variable that 
takes the mean of credit score upper and credit score lower. We will, when 
creating a histogram, start the x-axis at 300 to avoid lower values that are 
outliers or have omitted values. 

```{r}
loans$AvgCreditScore <- (loans$CreditScoreRangeLower + loans$CreditScoreRangeUpper) / 2
```

```{r}
ggplot(data = loans, aes(x = loans$AvgCreditScore), binwidth = 50)+
  geom_histogram()+
  scale_x_continuous(breaks = seq(250,850,50), limits = c(250,850))+
  ggtitle('Count of Loans by Average Credit Score')
```


From the above, we can see that most credit scores fall within 650 and 700. 
Let's bucket those credit scores to create a categorical variable for 
later analysis. 


```{r}
loans$csBucket <- cut(loans$AvgCreditScore, breaks = c(400, 500, 600, 700, 800, 
                                                       900))
ggplot(data = subset(loans, !is.na(csBucket)), aes(x=csBucket))+
  geom_bar()+
  ggtitle('Count of Loans by Credit Score Bucket')

```


When we look at the bucket bar chart below, we see there are a nearly equal 
amount of loans provided to average credit scores of between 600-700 and 
700-800, with very few happening below or above. 


# Univariate Analysis


### What is the structure of your dataset?

The structure of the dataset includes 113,937 observations of 81 variables, 
meaning there are 113,937 individual loans oberved. We have created two new 
variables to better examine credit score. avgCreditScore takes the average of 
the upper and lower credit score to take the mean for each individual borrower. 
The new variable CSBucket is a categorical variable that will help determine 
how credit may influence loans. 

### What is/are the main feature(s) of interest in your dataset?

The main features of interest for me in the dataset include: 
Prosper Score: What factors contribute most, or have the highest correlation, 
to establshing a Prosper rating. 

Also, is the LoanOriginalAmount for highly related to Income Range and 
credit score?  

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I am also interested in what levels of credit scores influence the eventual 
borrower APR. Are there distinct patterns between the credit worthiness of a 
borrower and whether they get a loan?

### Did you create any new variables from existing variables in the dataset?
Yes, I created the avgCreditScore variable to determine what number lies 
between the lower and upper credit score for each user. Then, I used the cut() 
function in order to bin the average credit scores per user. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form 
of the data? If so, why did you do this?

I had removed na's and blank values from a number of individual datapoints in 
order to better run analysis such as bar graphs. 


# Bivariate Plots Section

##Loan original amount by Prosper Rating
```{r echo=FALSE, Bivariate_Plots}
table(loans$ProsperRating..numeric.)
qplot(data = loans, y=loans$LoanOriginalAmount, x = loans$ProsperRating..Alpha., 
      geom='boxplot')+
  ggtitle('Loan Original Amount by Prosper Rating Alpha')
```


From the box plot above, we see that loans with Prosper Ratings of A, AA, and 
BB have much higher quartiles, outliers, and loan original amounts, which makes 
sense given that Prosper Ratings of C, D, E, HR are of lesser quality according 
to the Prosper website. 

Let's limit the ylim to further examine the median of the loan categories. 

```{r}
qplot(data = loans, y=loans$LoanOriginalAmount, x = loans$ProsperRating..Alpha., 
      geom='boxplot')+
  coord_cartesian(ylim = c(0,20000))+
  ggtitle('Loan Original Amount by Prosper Rating Alpha')
```


From the zoomed in data, we can see that AA loans have the highest median of any 
category at slightly above 10,000. Let's run a table to see what the 
median # actually is. 

```{r}
by(loans$LoanOriginalAmount, loans$ProsperRating..Alpha., summary)
```

The table above shows AA ratings with the median of 10,940, the highest of any 
category. 

##Loan original amount by Delinquencies Last 7 Years
We've examined the loan original amount from the Prosper Rating, let's see if 
borrower delinquencies have any effect on the loan original amount. 

```{r}
ggplot(data = loans, aes(x =loans$DelinquenciesLast7Years, 
                         y=loans$LoanOriginalAmount))+
  geom_point()+
  ggtitle('Loan Original Amount by Delinquencies Last 7 years')
```


The plot above shows some obvious levels to the data, as Propser seems to 
provide loans in 5,000 increments, however, we do see a slight pattern in which, 
when Delinquencies in the last 7 years rise, the total amount of a loan often 
falls. 

Let's add transparency and jitter in order to account for overplotting. 
```{r}
ggplot(data = loans, aes(x =loans$DelinquenciesLast7Years, 
                         y=loans$LoanOriginalAmount))+
  geom_point(alpha = 1/30, position = "jitter")+
  ggtitle('Loan Original Amount by Delinquencies Last 7 years')
```


The above chart, with it's dark line on the y-axis at 0, shows that Prosper 
very much likes to provide loans to those with little to no delinquencies in the 
past 7 years. There are a few points at 99 in the chart, which may be an anomoly 
or human error, and is worth further questioning of Prosper's data capture. 

##Loan original amount by Employment Status
Building on the picture we are beginning to see with regards to loan original 
amount, let's examine Employment status and loan original amount.

```{r}
summary(loans$EmploymentStatus)
```


In this data, I will create a new DF called loans_by_employmentstats as there 
are 2255 borrowers in which the status is blank, you can see this in the 
table above. 

```{r}
loans_by_employment <- subset(loans, loans$EmploymentStatus != "")
qplot(data = loans_by_employment, y=loans_by_employment$LoanOriginalAmount, x = loans_by_employment$EmploymentStatus, 
      geom='boxplot')+
  ggtitle('Loan Original Amount by Employment')
```


The boxplot above shows that employed Prosper borrowers, either through a 
company or self-employed, have higher median loan amounts than other categories, 
including Full-time and Part-time. This is odd, as employed persons are 
logically either employed full-time or part-time or self-employed. This is a 
question for Prosper as to what the classification methodology is. 

```{r}
qplot(data = loans_by_employment, y=loans_by_employment$LoanOriginalAmount, x = loans_by_employment$EmploymentStatus, 
      geom='boxplot')+
  coord_cartesian(ylim = c(0,20000))+
  geom_hline(yintercept = median(loans_by_employment$LoanOriginalAmount), 
             color = "green")+
  ggtitle('Loan Original Amount by Number of Investors and Employment Status')
```


By placing the overall median line on the graph, we see that only the categories
of "Employed" and "self-employed" actually have a higher median amount than the 
overall median, with the other categories falling behind. 

##Loan Original Amount by Estimated Return
We've got a sense of the borrowers, now let's take a look at the loan itself by 
examining the loan original amount and its estimated return. 
```{r}
ggplot(data = loans, aes(x=loans$EstimatedReturn, y=loans$LoanOriginalAmount))+
  geom_point(alpha = 1/20, position = 'jitter')+
  ggtitle('Loan Original Amount by Estimated Return')
```


The graph above, with slight transparency and jitter, show that the estimated 
return for most loans is falling around .05 and 1.5. Interestingly, there are a 
fair amount of loans that have negative returns, much more than those that have 
estimated returns of .2 or more. 

A Pearson correlation test below shows a negative correlation of -.28. I would 
have predicted greater correlation, however, the negative number suggests as the 
loan original amount increases, the estimated return decreases, a common 
phenomenmon.

```{r}
cor.test(loans$EstimatedReturn, loans$LoanOriginalAmount)
```



##Loan original amount by Investors
```{r}
mean(loans$Investors)
```

By taking the mean amount of investors per loan, which is calculated to around 
80, we can begin to see how many people it actually takes to get a Prosper loan 
fully-funed. A scatterplot can visualize this phenomeon. 

```{r}

ggplot(data = loans, aes(x = loans$Investors, y = loans$LoanOriginalAmount))+
  geom_point(alpha = 1/10)+
  geom_vline(xintercept = mean(loans$Investors), color = "green")+
  ggtitle('Loan Original Amount by Number of Investors')
```


The scatter plot above shows that, as loan amounts grow, so do the total amount 
of investors needed. The mean of just above 80 shows that the majority of loans 
are falling at a similar price level, thus a similar amount of investors are 
needed for analogous loans. 

A correlation below shows a Pearson score of .38, lesser than what I expected. 

```{r}
cor.test(loans$Investors, loans$LoanOriginalAmount)
```

##Loan original amount by Credit Score

We will, finally, use our newly binned data to examine loan original amount by 
credit score bucket. This will provide a nice picture as to which credit score 
buckets have the highest median Loan Original amount. 

```{r}
qplot(data = loans, y=LoanOriginalAmount, x = csBucket, 
      geom='boxplot')+
  geom_hline(yintercept = median(loans$LoanOriginalAmount), color = "green")+
  ggtitle('Loan Original Amount by Credit Score Bucket')
```


Unsurprsiingly, the buckets of 700-800 and 800-900 both have medians that exceed 
the overall median. It is interesting to note that 600-700 credit scores have a 
median Loan Original Amount less than the overall median, an indication credit 
scores in this area are not deemed as worthy for a loan. 


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Most interestingly, I discovered that a credit score of 600-700 has a 
substantially lower median loan original amount than credit scores of 700-800. 
This was surprising to me, as I suspected a slight decrease but, personally, 
have always considered a credit score of 600-700 to be somewhat trustworthy. 

Also, the total amount of investors per loan was quite intriguing, as there 
were substantial outliers from the mean of 80 per loan. We will further examine 
this by exploring categories that may influence that in the multi-variate 
analysis portion.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
Yes, the relationship between employment status was particularly interesting, 
as "full time" employees have a signficiantly lower median loan amount than 
those simply categorized as "employed". I suspect there may be an error in 
Prosper's classification system. 

### What was the strongest relationship you found?
The relationship between total investors and dollar amount had a Pearson score 
of .38, the strongest relationship I uncovered.

Interestingly, loan original amount and estimated return were negatively 
correlated, showing that as a loan amount is greater, estimated returns are 
generally expected to fail. 

# Multivariate Plots Section

##Examining loan original amound and borrower APR by Credit Score Bucket.

```{r echo=FALSE, Multivariate_Plots}
ggplot(data = subset(loans,!is.na(loans$csBucket)), 
       aes(x = BorrowerAPR, y = LoanOriginalAmount))+
  geom_point(alpha = 1/2, aes(color = csBucket))+
  ggtitle('Loan Original Amount by Borrower APR and Credit Score Bucket')
```


The above scatterplot shows that csBucket 800-900 are largely consolidated at a 
little less than .1 for their Borrower APR, despite loan amount. This shows that 
Prosper generally loans funds to those with good credit at various levels of 
dollar amount.

You can see distinct, almost vertical patterns of increased BorrowerAPR based 
on their credit score. Lower credit scores from 400-600 are largely consolidated 
lower on the y-axis at varying APRs, showing that credit scores that are lesser 
do not get loaned as large amounts as other credit scores. 



##Examining loan original amount and estimated return by Credit Score Bucket.

```{r}
ggplot(data = subset(loans,!is.na(loans$csBucket)), 
       aes(x = EstimatedReturn, y = LoanOriginalAmount))+
  geom_point(aes(color = csBucket))+
  ggtitle('Loan Original Amount by Estimated Return and Credit Score Bucket')
```


Similarly to our analysis on borrower apr, estimated returns are largely 
attributed to certain credit scores. The highest end of credit scores (800-900), 
have a distinct Estimated Return of around .05 percent, with 700-800 categories 
at around .075 to .1, and 600-700 following thereafter with growing estimated 
returns. This could be because those with higher credit scores pay their loans 
off quicker, at lower rates. 

Interestingly, there are a bevy of loans at the 600-700 credit score level that 
do not pay off loans that are less than 10,000 in total value. Even though they 
have credit scores similarly to those borrowing at higher dollar amounts, they 
were unable to pay their comparable smaller debts. 

##Loan Original Amount, # of Investors, Credit Score

```{r}
ggplot(data = subset(loans,!is.na(loans$csBucket)), 
       aes(x = Investors, y = LoanOriginalAmount))+
  geom_point(alpha = 1/3, aes(color = csBucket))+
  geom_vline(xintercept = median(loans$Investors))+
  ggtitle('Loan Original Amount by Number of Investors and Credit Score Bucket')
```


The graph above shows that the total amount of investors increases as the 
loan original amount increases, and invariably, the higher credit score users 
are more adept at getting additional investors. In the scatterplot, you hardly
see any 400-700 credit score borrowers beyond the overall median line.

Interestingly, the higher end credit score borrowers of 800-900 generally have 
a # of investors per loan that is higher than the median and higher than there 
lesser counterparts, even twhen the loan original amount is the same, this means 
more individuals are investing smaller amounts per individual loan. 


##Loan Original Amount, Loan Investors, Prosper Rating

We run the tables below to view which Prosper Rating has the most loans and the 
highest loan amounts, we find that a score of 7 receives the highest median loan 
and a score of 1-2 have the lowest medians. This may be an indication of the 
loan worthiness of the borrowers - thus, we assue 7 to be the best loan for 
chance of repayment and 1 being the worse. 

```{r}
table(loans$ProsperRating..numeric.)
```

```{r}
by(loans$LoanOriginalAmount, loans$ProsperRating..numeric., summary)
```

```{r}
ggplot(data = loans, aes(x = Investors, y = LoanOriginalAmount))+
  geom_point(alpha = 1/5, position = 'jitter', 
             aes(color = factor(loans$ProsperRating..numeric.)))+
  geom_vline(xintercept = median(loans$Investors))+
  scale_color_brewer(palette = "RdYlGn")+
  ggtitle('Loan Original Amount by Number of Investors and 
          Numeric Prosper Rating')
```

The graph above shows that, as the number of investors and the loan original 
amount increases, more high-quality loans of Prosper Ratings of 4-5-6-7 are 
prevalent. 

##Delinquencies 7 Years Loan Investors, Prosper Rating Numeric
```{r}
ggplot(data = loans_by_employment, 
       aes(x = DelinquenciesLast7Years, y = LoanOriginalAmount))+
  geom_point(alpha = 1/2, position = 'jitter', 
             aes(color = loans_by_employment$EmploymentStatus))+
  geom_vline(xintercept = 
               mean(loans_by_employment$DelinquenciesLast7Years), 
             color = 'black')+
  ggtitle('Loan Original Amount by Delinquencies Last 7 Years and 
          Employment Status In U.S. Dollars (USD)')
```

When looking at the loan original amount and delinquencies the last 7 years, 
it is no surprise that the majority of points on the graph are those that are 
employed, even when transparency is added for overplotting. With the "employed" 
category as "red", you will notice the prevalence of red dots as the loan amount 
increases, and the lack of red dots as the delinquencies the last 7 years 
increase below $5,000 loan amounts. 

This means that those who are not categorized as "employed" in the Prosper 
system have higher delinquencies in the past 7 years, and do not get the same 
level of loan as those who are employed. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In this section, we wanted to look closer at what categories may be determining 
loan amounts, the amount of investors, and the delinquencies in the last 7 
years. We noticed that those who were not categorized as "employed" saw 
themselves receiving loans of roughly less than 5,000. Those same individuals 
were likely to have a higher rate of delinquencies in the last 7 years than 
those categorized as "employed", and their loan amounts were lesser than their 
employed counterparts even if the delinquencies in the last 7 years were 
the same. 

We also discovered that Prospers numeric rating runs from 1-7, with 7 being the 
most trustworthy rating.

### Were there any interesting or surprising interactions between features?

I was overall surprised to see that the number of investors increases as credit 
score and prosper rating increases. It was my inclination that if a loan were to 
be deemed safe or less risky, more investors would provide higher individual 
amounts towards that loan, thus, lowering the total amount of investors per 
loan.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

We did not create any models with this dataset. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data = loans_by_employment, 
       aes(x = DelinquenciesLast7Years, y = LoanOriginalAmount))+
  geom_point(alpha = 1/2, position = 'jitter', 
             aes(color = loans_by_employment$EmploymentStatus))+
  geom_smooth(method ="lm", aes(color = loans_by_employment$EmploymentStatus))+
  ggtitle('Loan Original Amount In U.S. Dollars (USD) by Delinquencies Last 7 
Years and Employment Status')+
  xlab('Number of Delinquencies Last 7 Years')+
  ylab('Loan Original Amount In U.S. Dollars (USD)')+
  labs(colour="Employment Status")+
  scale_color_brewer(type='seq', guide=guide_legend(title='Employment Status'), 
                     palette = "Set1")
```


### Description One
The plot above shows the Loan Original Amount on the y-axis, the number of 
delinquencies in the last 7 years on the x-axis, with each plot indicating an 
employment status as a categorical variable.

This graph shows to me that, should you be employed, you have a marginal chance 
of receiving a loan from Prosper. Though the large majority of loans are at 0 
delinquencies, it is important to notice that there are a great number of loans 
provided to people who have be delinquent dozens of times. You are more likely 
to receive a greater loan amount, however, if your employment status 
is "employed"

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data = subset(loans,!is.na(loans$csBucket)), 
       aes(x = Investors, y = LoanOriginalAmount))+
  geom_point(alpha = 1/3, aes(color = csBucket))+
  ggtitle('Loan Original Amount In U.S. Dollars (USD) by Number of Investors 
and Credit Score Bucket')+
  xlab('Number of Investors')+
  ylab('Loan Original Amount In U.S. Dollars (USD)')+
  labs(colour="Credit Score")+
  geom_smooth(method="lm", se=FALSE, aes(color = csBucket))+
  scale_color_brewer(type='seq', guide=guide_legend(title='Employment Status'), 
                     palette = "Set1")
```


### Description Two
The graph above shows the number of investors per loan on the x-axis, the total 
loan original amount on the y-axis, with the categorical variable of credit 
score bucket. This plot shows that, as the loan amount increases, the number of 
investors increases as well.

We already know from prior graphs that higher loan amounts are more likely to be 
people employed with no delinquencies on their account, couple that with higher 
credit scores and higher loan amounts are generally more safe. We see that, as 
loan amounts increase, the number of investors per loan also increases. 

This, to me, is a strange phenomenmon, as I would have guessed that, if a loan 
were to be less risky, investors would invest greater amounts towards that loan, 
thus limiting the total amount of investors per loan. 

The plot shows borrowers with credit scores between 700-900 are generally 
receiving more investors at higher loan amounts than other credit scores. 

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = subset(loans,!is.na(loans$csBucket)), 
       aes(x = EstimatedReturn, y = LoanOriginalAmount))+
  geom_point(aes(color = csBucket))+
   geom_smooth(method="lm", se=FALSE, aes(color = csBucket))+
  ggtitle('Loan Original Amount In U.S. Dollars (USD) by Estimated Return and 
          Credit Score Bucket')+
  xlab('Estimated Return on Investment')+
  ylab('Loan Original Amount In U.S. Dollars (USD)')+
  labs(colour="Credit Score")+
  scale_color_brewer(type='seq', guide=guide_legend(title='Credit Score'), 
                     palette = "Set3")
```


### Description Three
The above plot shows the estimated return per loan on the x-axis, with the loan 
original amount on the y-axis and the credit score bucket as the categorical 
variable.

This plot shows that estimated returns will generally fall between .05 and .15, 
are nearly normally distributed, and will be somewhat dependent on the credit 
score of the borrower. Borrowers with higher credit scores are shown to provide 
less returns to investors, likely due to lower interest rates and ability to 
pay back a loan quickly. 

The plot also shows that estimated returns for higher credit score borrowers are 
largely the same regardless of the loan amount, showing that loan amount does 
not really affect return rate, and vice versa.
------

# Reflection

The Prosper data was quite large and there were a number of variables that were 
omitted from this analysis simply due to time and expectations from the course. 
With greater time, one could examine the areas such as investment from friends 
to see if certain credit score users or certain Prosper loans are more likely to 
get loans from their friends.

It was interesting to note that there are a higher number of investors per loan, 
particularly when the credit score of the borrower is high. This goes against my 
intution that single or a handful of investors would want to buy out a loan 
completely. This potentially may show some skepticism of the Prosper platform.

I did not conduct any analysis of loan variables over time, as my questions were 
more focused on what affects a loan amount and the amount of investors investing 
in said loan. It may be interesting to explore whether estimated returns lowered 
or expanded overtime for Prosper loans. 

